---
title: 'Abstract 3: Report'
author: "Lucas Veras"
date: "2019-01-21"
output: 
  html_document:
    keep_md: true
    toc: true
    code_folding: show
editor_options: 
  chunk_output_type: console
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r prep, echo = FALSE}
library(knitr)
library(gridExtra)
library(tidyverse)
library(nlme)
source("R/cvMixedModel.R")
source("R/accuracyIndices.R")

hip.base <- read_csv("data/regressions_comparison/Hip_sec_base.csv")
hip.acc  <- read_csv("data/regressions_comparison/Hip_sec_acc_metrics.csv")

hip <- hip.base %>% 
  left_join(hip.acc, by = c("ID", "speed")) %>% 
  dplyr::select(-c(Evaluation, Date))
```

## Overview

Accelerometry is recognized as a valid method to measure energy expenditure (EE). However, high prediction accuracy is only expected for people with similar characteristics to those of the utilized in the prediction model samples, such as body composition. Currently, prediction models based on raw acceleration metrics have been only developed for non-obese people, and their accuracy is unknown on severe obese people. Therefore, the aim of this study was to develop an equation to predict EE in class II-III obese people, determine its prediction accuracy and compare it with those equations developed for non-obese people.

## Data analysis

### Sample descriptives
```{r sample_descriptives, echo = FALSE}
samp.desc <- read_csv("data/regressions_comparison/sample_descriptives_data.csv")
descriptives <- summarise(
  .data = samp.desc,
  age_mean    = round(mean(age), digits = 2),
  age_sd      = round(sd(age), digits = 2),
  weight_mean = round(mean(weight_kg), digits = 2),
  weight_sd   = round(sd(weight_kg), digits = 2),
  height_mean = round(mean(height_cm), digits = 2),
  height_sd   = round(sd(height_cm), digits = 2),
  BMI_mean    = round(mean(BMI_kgm2), digits = 2),
  BMI_sd      = round(sd(BMI_kgm2), digits = 2),
  fat_mean    = round(mean(body_fat), digits = 2),
  fat_sd      = round(sd(body_fat), digits = 2)
)
sex <- table(samp.desc$sex)
```

`r nrow(samp.desc)` class II-II obese adults (`r sex[[2]]` males, `r sex[[1]]` females; age: `r descriptives$age_mean`±`r descriptives$age_sd`yrs; weight: `r descriptives$weight_mean`±`r descriptives$weight_sd`kg; height: `r descriptives$height_mean`±`r descriptives$height_sd`cm; body fat: `r descriptives$fat_mean`±`r descriptives$fat_sd`%; BMI: `r descriptives$BMI_mean`±`r descriptives$BMI_sd`kg^.^m^-1^)

### Mixed models

Two linear mixed models were build to predict VO~2~/kg from accelerometer metrics (MAD and ENMO). Final models had linear and quadratic terms of accelerometer metrics and age as fixed effect and subjects as random effect. Models formulas and summaries can be seen below.

```{r mixed_models}
model.MAD <- lme(
  fixed = VO2.kg ~ MAD + I(MAD^2) + Age,
  random = ~ 1 | ID,
  method = "ML",
  correlation = corAR1(),
  data = hip,
  na.action = na.omit
)

model.ENMO <- lme(
  fixed = VO2.kg ~ ENMO + I(ENMO^2) + Age,
  random = ~ 1 | ID,
  method = "ML",
  correlation = corAR1(),
  data = hip,
  na.action = na.omit
)

summary(model.MAD)
summary(model.ENMO)
```

After that, both models were cross-validated using the leave-one-out approach. Leave-one-out cross validation data were used to calculate indices of accuracy for the non-filtered accelerometer metrics.

```{r LOOCV}
fix.eff   <- VO2.kg ~ MAD + I(MAD^2) + Age
rand.eff  <- ~ 1 | ID
MAD.LOOCV <- do.call(rbind, (lapply(unique(hip$ID), cvMixedModel, df = hip)))

fix.eff    <- VO2.kg ~ ENMO + I(ENMO^2) + Age
rand.eff   <- ~ 1 | ID
ENMO.LOOCV <- do.call(rbind, (lapply(unique(hip$ID), cvMixedModel, df = hip)))

MAD.LOOCV  <- select(MAD.LOOCV, ID, speed, VO2.kg, VO2.kg_predicted)
ENMO.LOOCV <- select(ENMO.LOOCV, ID, speed, VO2.kg, VO2.kg_predicted)
```

### Prediction models developed for non-obese people

Our accelerometer metrics data was used to predict VO~2~/kg with regression equations developed for non-obese people. Prediction models utilized as reference were the Vähä-Ypyä equation (1), that used MAD metric, and the Hildebrand equation (2), that used ENMO metric

```{r non_obese_models}
# MAD
# Vaha-Ypya H, et al. PloS one. 2015;10(8):e0134813
# (prediction model based on linear regression for walking only)
MAD.VahY <- hip %>% 
  select(ID, speed, VO2.kg, MAD) %>% 
  na.omit()
MAD.VahY$VO2.kg_predicted <- NA
for (i in 1:nrow(MAD.VahY)) {
  MAD.VahY$VO2.kg_predicted[i] <- 7.92 + 0.0331 * MAD.VahY$MAD[i]
}

# ENMO
# Hildebrand M, et al. Med Sci Sports Exerc. 2014 Sep;46(9):1816-24
# (equation for adults, actigraph at hip)
ENMO.Hild <- hip %>% 
  select(ID, speed, VO2.kg, ENMO) %>% 
  na.omit()
ENMO.Hild$VO2.kg_predicted <- NA
for (i in 1:nrow(ENMO.Hild)) {
  ENMO.Hild$VO2.kg_predicted[i] <- 0.0554 * ENMO.Hild$ENMO[i] + 6.67
}
```

### Indices of accuracy

Indices of accuracy were calculated comparing VO~2~/kg measured by indirect calorimetry with predicted by the equations. Values of theses indices are shown below.
```{r accuracy_indices}
MAD.LOOCV.accuracy  <- accuracyIndices(MAD.LOOCV, "VO2.kg_predicted", "VO2.kg")
MAD.VahY.accuracy   <- accuracyIndices(MAD.VahY, "VO2.kg_predicted", "VO2.kg")
ENMO.LOOCV.accuracy <- accuracyIndices(ENMO.LOOCV, "VO2.kg_predicted", "VO2.kg")
ENMO.Hild.accuracy  <- accuracyIndices(ENMO.Hild, "VO2.kg_predicted", "VO2.kg")

accuracy.all <- rbind(MAD.LOOCV.accuracy, MAD.VahY.accuracy, ENMO.LOOCV.accuracy, ENMO.Hild.accuracy)
accuracy.all$metric <- c(rep("MAD", 2), rep("ENMO", 2))
accuracy.all$equation <- c("Our", "Vaha-Ypya", "Our", "Hildebrand")
accuracy.all <- accuracy.all[, c(15, 16, 1:14)]

kable(accuracy.all)
```

Predictors included in our equations were age and one of the accelerometer metrics. Bland-Altman plots from our equations presented good agreement between measured and predicted VO2, with an irrelevant bias (between -0.02 and 0.01 ml∙kg−1∙min−1), contrarily to reference equations that revealed a substantial bias (< -3.00 ml∙kg−1∙min−1). Our equation that utilized MAD metric revealed a mean absolute percent error of 13.72%, while with Vähä-Ypyä equation was 57.10%. The same pattern occurred with equations that utilized ENMO metric, with our equation presenting a mean absolute percent error of 17.29%, while with Hildebrand equation was 51.57%. 

### Bland-Altman plots
 
Bland-Altman plots for predictions based on MAD metric using our and Vähä-Ypyä equations are shown in Figure 1 and Figure 2, respectively.
 
```{r MAD_BAplots, dev = 'svg', fig.width = 12, fig.height = 6}
# MAD - our equation
BAplot.MAD.our <- ggplot(data = MAD.LOOCV, aes(x = ((VO2.kg + VO2.kg_predicted) / 2), y = (VO2.kg - VO2.kg_predicted))) +
  geom_point() +
  geom_hline(yintercept = mean(MAD.LOOCV$VO2.kg - MAD.LOOCV$VO2.kg_predicted)) +
  geom_hline(
    yintercept = mean(MAD.LOOCV$VO2.kg - MAD.LOOCV$VO2.kg_predicted) + 
                 (1.96 * sd(MAD.LOOCV$VO2.kg - MAD.LOOCV$VO2.kg_predicted)),
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = mean(MAD.LOOCV$VO2.kg - MAD.LOOCV$VO2.kg_predicted) - 
      (1.96 * sd(MAD.LOOCV$VO2.kg - MAD.LOOCV$VO2.kg_predicted)),
    linetype = "dotted"
  ) +
  scale_y_continuous(limits = c(-6, 6), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0, 20), expand = c(0, 0)) +
  theme_classic() +
  theme(
    axis.line = element_line(size = 1, colour = "Black"),
    axis.ticks = element_line(size = 1, colour = "Black"),
    axis.ticks.length = unit(2, "mm"),
    plot.title = element_text(face = "bold")
  ) +
  labs(
    title = "Figure 1",
    x = bquote("Mean of Measured and Predicted" ~ VO[2] ~ (ml^. ~ kg^-1~ . ~ min^-1)),
    y = bquote("Difference of Measured and Predicted" ~ VO[2] ~ (ml^. ~ kg^-1~ . ~ min^-1))
  )

# MAD - Vaha-Ypya equation
BAplot.MAD.VahY <- ggplot(data = MAD.VahY, aes(x = ((VO2.kg + VO2.kg_predicted) / 2), y = (VO2.kg - VO2.kg_predicted))) +
  geom_point() +
  geom_hline(yintercept = mean(MAD.VahY$VO2.kg - MAD.VahY$VO2.kg_predicted)) +
  geom_hline(
    yintercept = mean(MAD.VahY$VO2.kg - MAD.VahY$VO2.kg_predicted) + 
      (1.96 * sd(MAD.VahY$VO2.kg - MAD.VahY$VO2.kg_predicted)),
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = mean(MAD.VahY$VO2.kg - MAD.VahY$VO2.kg_predicted) - 
      (1.96 * sd(MAD.VahY$VO2.kg - MAD.VahY$VO2.kg_predicted)),
    linetype = "dotted"
  ) +
  scale_y_continuous(limits = c(-12, 4), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0, 25), expand = c(0, 0)) +
  theme_classic() +
  theme(
    axis.line = element_line(size = 1, colour = "Black"),
    axis.ticks = element_line(size = 1, colour = "Black"),
    axis.ticks.length = unit(2, "mm"),
    plot.title = element_text(face = "bold")
  ) +
  labs(
    title = "Figure 2",
    x = bquote("Mean of Measured and Predicted" ~ VO[2] ~ (ml^. ~ kg^-1~ . ~ min^-1)),
    y = bquote("Difference of Measured and Predicted" ~ VO[2] ~ (ml^. ~ kg^-1~ . ~ min^-1))
  )

grid.arrange(BAplot.MAD.our, BAplot.MAD.VahY, ncol = 2)
```

Bland-Altman plots for predictions based on ENMO metric using our and Hildebrand equations are shown in Figure 3 and Figure 4, respectively.

```{r ENMO_BAplots, dev = 'svg', fig.width = 12, fig.height = 6}
# ENMO - our equation
BAplot.ENMO.our <- ggplot(data = ENMO.LOOCV, aes(x = ((VO2.kg + VO2.kg_predicted) / 2), y = (VO2.kg - VO2.kg_predicted))) +
  geom_point() +
  geom_hline(yintercept = mean(ENMO.LOOCV$VO2.kg - ENMO.LOOCV$VO2.kg_predicted)) +
  geom_hline(
    yintercept = mean(ENMO.LOOCV$VO2.kg - ENMO.LOOCV$VO2.kg_predicted) + 
      (1.96 * sd(ENMO.LOOCV$VO2.kg - ENMO.LOOCV$VO2.kg_predicted)),
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = mean(ENMO.LOOCV$VO2.kg - ENMO.LOOCV$VO2.kg_predicted) - 
      (1.96 * sd(ENMO.LOOCV$VO2.kg - ENMO.LOOCV$VO2.kg_predicted)),
    linetype = "dotted"
  ) +
  scale_y_continuous(limits = c(-6, 6), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0, 20), expand = c(0, 0)) +
  theme_classic() +
  theme(
    axis.line = element_line(size = 1, colour = "Black"),
    axis.ticks = element_line(size = 1, colour = "Black"),
    axis.ticks.length = unit(2, "mm"),
    plot.title = element_text(face = "bold")
  ) +
  labs(
    title = "Figure 3",
    x = bquote("Mean of Measured and Predicted" ~ VO[2] ~ (ml^. ~ kg^-1~ . ~ min^-1)),
    y = bquote("Difference of Measured and Predicted" ~ VO[2] ~ (ml^. ~ kg^-1~ . ~ min^-1))
  )

# ENMO - Hildebrand equation
BAplot.ENMO.Hild <- ggplot(data = ENMO.Hild, aes(x = ((VO2.kg + VO2.kg_predicted) / 2), y = (VO2.kg - VO2.kg_predicted))) +
  geom_point() +
  geom_hline(yintercept = mean(ENMO.Hild$VO2.kg - ENMO.Hild$VO2.kg_predicted)) +
  geom_hline(
    yintercept = mean(ENMO.Hild$VO2.kg - ENMO.Hild$VO2.kg_predicted) + 
      (1.96 * sd(ENMO.Hild$VO2.kg - ENMO.Hild$VO2.kg_predicted)),
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = mean(ENMO.Hild$VO2.kg - ENMO.Hild$VO2.kg_predicted) - 
      (1.96 * sd(ENMO.Hild$VO2.kg - ENMO.Hild$VO2.kg_predicted)),
    linetype = "dotted"
  ) +
  scale_y_continuous(limits = c(-10, 2), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0, 25), expand = c(0, 0)) +
  theme_classic() +
  theme(
    axis.line = element_line(size = 1, colour = "Black"),
    axis.ticks = element_line(size = 1, colour = "Black"),
    axis.ticks.length = unit(2, "mm"),
    plot.title = element_text(face = "bold")
  ) +
  labs(
    title = "Figure 4",
    x = bquote("Mean of Measured and Predicted" ~ VO[2] ~ (ml^. ~ kg^-1~ . ~ min^-1)),
    y = bquote("Difference of Measured and Predicted" ~ VO[2] ~ (ml^. ~ kg^-1~ . ~ min^-1))
  )

grid.arrange(BAplot.ENMO.our, BAplot.ENMO.Hild, ncol = 2)
```

## R session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```